Imprecise Datapath Design: An Overclocking Approach

### 2013-06-21 TRETS2013/TRETS_overclocking.tex
TRETS revised by George, words and sentences

% Title portion
\title{Imprecise Datapath Design: An Overclocking Approach}
\author{KAN SHI, DAVID BOLAND\\
and GEORGE A CONSTANTINIDES
\affil{Imperial College London}
}

\begin{abstract}
As process scaling introduces significant performance variations, much more conservative timing margins will have to be applied to avoid timing violations. Potentially, releasing the tight accuracy requirement would offer greater freedom to create a design with better performance or energy efficiency. In this paper, we compare two different approaches that could trade accuracy for performance. One is the traditional approach where the precision used in the datapath is limited to meet a target latency. The other is a proposed new approach which simply allows the datapath to operate without timing closure. We demonstrate analytically and experimentally that on average our approach obtains either smaller errors or faster operating frequencies in comparison to the traditional approach. This is because the worst case caused by timing violations only happens rarely, while the precision loss results in errors to most data. We also show that for basic arithmetic operations such as addition, applying our approach to simple operators (i.e. ripple carry adders) can achieve better accuracy or performance than using the traditional method to both original operators and intrinsically faster operators (i.e. carry select adders) which consumes the same silicon area.
\end{abstract}

@@

\section{Introduction}
With the continuous scaling of CMOS technology to the nanometer regime, circuit performance has increased tremendously over the past decades. However, the drastic variations introduced by higher integration densities is anticipated to be the major obstacle when designing reliable, high performance circuits. Currently, hardware designers tend to employ conservative safety margins to ensure a uniform functionality across a variety of possible working environments. Nevertheless, continuing with this approach could become very costly in the short term future. This is because the highly scaled CMOS devices would inevitably exhibit probabilistic or statistical behavior, meaning that we can no longer expect a uniform circuit performance. Furthermore, covering all possible worst cases would become increasingly difficult, expensive and result in large yield loss because of the variation.

Specifically, a large volume of current studies has demonstrated that, significant performance gains can be achieved by FPGA-based accelerators over software designs across a wide range of applications~\cite{UnderwoodFPGAvsCPU,David_MINRES}. However, one of the major factors that limits the performance of these accelerators is that they typically run at much lower clock frequencies than general purpose processors (GPPs) or graphics processing units (GPUs).

In order to boost the operating frequency of a datapath, the standard approaches are either to heavily pipeline the design or to reduce the precision throughout the datapath. For the former method, it should be noted that pipelining will not tend to reduce the latency of the datapath, whilst the latency in terms of clock cycles will actually increase. As a result, this method will not be applicable to many embedded applications, which typically have strict latency requirements, or in any datapath containing feedback where C-slow retiming is inappropriate. For the second approach, reducing the datapath precision would reduce the latency at the cost of introducing quantization errors into the design. This has inspired research into exploiting the potential benefits of using the minimum precision necessary to satisfy a design specification~\cite{GeorgeDT11}. A brief review of existing approaches in this area will be discussed in Section~\ref{Section_Background_WL}.

Unfortunately, neither of these conventional approaches tends to remove the conservative safety margin. In recent years, we have seen a growth of research that explores the potential power or performance benefits that can be obtained when relaxing the variation-induced guard-bands~\cite{NewBox2004}. The detailed review of this research field will be presented in Section~\ref{Section_Background_Imprecise}.

In this paper, we describe an alternative circuit design methodology when considering trade-offs between accuracy, performance and silicon area. We suggest that for certain applications it is beneficial to move away from the traditional model of creating a conservative design that is guaranteed to avoid timing violations. Instead, it may be preferable to create a design in which timing violations may occur, under the knowledge that they only occur rarely because specific input patterns are required to generate the worst case errors. This paper elaborates on the prior work [xxx], by incorporating silicon area as another evaluation metric. This enables us to find the optimum design choice of the basic arithmetic operators (i.e. binary adders) under various design constraints. A summary of the main contributions of this work are as follows:

\begin{itemize}
	\item Detailed modeling methods of probabilistic errors generated in basic arithmetic primitives from two design scenarios to trade accuracy for performance: overclocking the datapath or truncating precisions that used throughout the datpath,
	\item Exploring the accuracy-performance-area trade-offs for key arithmetic operators to illustrate the optimum architecture for a certain arithmetic function under a given requirement of silicon area,
	\item Demonstration of the proposed findings analytically and experimentially from FPGA implementations that allowing rare timing violations to happen leads to an reduction of average errors or an improvement of performance over trancating the datapath precisions to meet timing.
\end{itemize}

@@

\section{Background}\label{Section_Background}
\subsection{Word-length Optimization}\label{Section_Background_WL}
There exists a significant amount of work demonstrating that optimizing the precision used throughout a datapath would bring a substantial benefits on clock speed, silicon area and power consumption. There are two main research directions in this area. The first discusses methods to analyze the errors caused by the usage of finite precision and compute bounds on the worst-case errors seen at the output of the design. Related techniques include the simulation-based approach~\cite{SungWLop_Simulation} and the analytical-based approach such as interval arithmetic~\cite{moore1966IA}, affine arithmetic~\cite{de2004affine}, Satisfiability-Modulo Theories~\cite{SAT2010} and polynomial representations~\cite{Boland2011TCAD}. A detailed summary of these methods can be found in~\cite{GeorgeDT11,lee2006accuracy}. In the second research field, these tools are utilized to choose the minimum precision within a datapath to satisfy a required accuracy specification~\cite{RoldaoWLop_Simulation}. In this work, we speculate and employ the maximum word-length in the datapath under a given timing constraint while ensuring no timing errors will occur. The truncation errors derived in this process are then compared to the overclocking errors generated in our proposed new scenario.

\subsection{Imprecise Circuit Design Methodology}\label{Section_Background_Imprecise}
It is worth noting that the choice of precision is not the only source of error when designing a datapath. Recently we have seen a growth of parallel streams of research which aim at exploring alternative methods to trade accuracy for design efficiency. This strand of research is motivated by the fact that extra benefits of manufacturing, test, power and timing can be obtained if the tight requirement of absolute correctness is released for devices and interconnects, as pointed out by the International Technology Roadmap for Semiconductors (ITRS) in 2007~\cite{semiconductor2007international}. This topic is expected to be of growing importance in the future, because we will face new design challenges as the technology scales further.

According to the existing literatures, one way to operating circuits beyond the deterministic region is to relax the design constraints and the guard-bands that are conventionally used to avoid the worst cases. A series of work known as ``Better-Than-Worst-Case (BTWC) Design'' introduced a universal framework to push the circuit performance to its limits~\cite{BetterThanWS2005}. In general, a BTWC design is composed of cores and checkers. The cores are operated with high performance by eliminating the guard-bands, meanwhile the possible timing errors are diagnosed by the checkers. Optionally the system can be recovered at the observation of errors. As an exemplary design, the Razor project~\cite{Razor2003,Razor2004} was proposed to shave the conservative timing margins by overscaling the supply voltage and clock frequency, while monitoring the output error rates by utilizing a self-checking circuit. This work demonstrated that the benefits brought by removing the safe margin outweigh the cost of monitoring and recovering from errors. For example, $22\%$ or over $30\%$ power consumption can be saved with $\sim0.01\%$ or $\sim1\%$ error rates at the output, respectively. Related work also includes a similar frequency overscaling technique by operating circuits slightly slower than the critical path delay with dedicated checker circuits to ensure that timing errors will not occur~\cite{TEAtime2004}, or developing timing analysis tools that decide the optimum operating frequencies in the non-deterministic region due to process variation~\cite{ProbabilisticSTA}. However, the major problem of the BTWC work is the extra cost introduced by implementing the checker circuits.
%of silicon area due to the implementation of the checker circuits.

Another field of study has shown that errors can be potentially tolerated in various applications, e.g. DSP applications concerning human perceptions. In this case, even greater performance or power benefits can be achieved, as the design overhead of the checkers can be eliminated. Current research in this area focuses on designing probabilistic or imprecise circuits both at a software level and a hardware level. For instance at the programming language level, a tool was proposed~\cite{EnerJ2011Uwash,Truffle2012Uwash} to divide the program into the precise parts and the approximate parts, which are mapped to different hardware with different speed-grades, supply voltages etc., respectively. This technique enables a relatively high service quality, in the meantime energy reduction can be obtained due to approximation. Similar ideas have also been applied on hardware directly. As an example, Palem et~al.~\cite{NonUniformScaling} described a non-uniform voltage scaling technique for the ripple carry adder. In this study, multiple voltage regions are employed for different bits along a carry chain. That is, higher voltage would be applied for computations generating most significant bits, and vice versa. Nevertheless, utilizing several voltage regions within a ripple carry adder will rarely be practical in real situations. Furthermore, the overhead of applying this technique is not addressed.

While the aforementioned literatures take advantage of the fact that only specific input patterns could cause timing errors, research on imprecise architectures take this one step further by designing simplified circuits for peformance and/or energy efficiency. For example, Lu et~al. proposed a ``shrinking'' datapath that can be utilized to mimic and speculate the original logic functions~\cite{IntelSpeeding}. Kulkarni et.al. described an underdesigned $2\times2$ multiplier unit, of which the worst case was replaced by a normal case based on the straight-forward Karnaugh-Map analysis~\cite{Undersigned2x2multiplier}. In both cases, reduction of area and power consumption can be achieved with the cost of accuracy. In addition, Gupta et~al. developed approximate adders at the transistor level and compared the energy efficiency of their proposed architectures over truncation of input word-length of conventional structures~\cite{Gupta2013TransCADICS}. However, we should note that using imprecise architectures means correct results would never be obtained for certain input patterns. Another limitation is that the link between the probability of output correctness and energy savings or performance improvements are rarely analyzed from current research. In addition, these techniques cannot be directly applied onto existing hardware platforms such as FPGAs.

In this work, we propose an alternative methodology to remove the limitations of those existing approaches. It is argued in this paper that our method can be easily applied onto existing hardware arithmetic operators such as RCAs and CCMs with almost no hardware overhead. We support this hypothesis analytically by presenting theoretical probabilistic error models and experimentally using FPGAs.

\section{Ripple Carry Adder}\label{section_RCA}
\subsection{Adder Structures in FPGAs}
Adders serve as a key building block for arithmetic operations. In general, the ripple carry adder (RCA) is the most straightforward and widely used adder structure. As such, the philosophy of our approach is first exemplified with the analysis of a RCA. We later describe how this methodology can be extended to other arithmetic operators such as CSAs and CCMs, which are commonly used in DSP applicationsn and numerical algorithms, in Section~\ref{Section_CSA} and Section~\ref{section_CCM} respectively.

@@

While the fast carry logic reduces the time of each individual carry-propagation delay, the overall delay of carry-propagation will eventually overwhelm the delay of sum generation of each LUT with increasing operand word-lengths. We perform our initial analysis based on the assumption that the carry propagation delay of each FA is a constant value $\mu_c$, which is a combination of logic delay and routing delay within an FPGA. Hence the critical path delay of the RCA is given by~(\ref{Eq:RCA_delay}), as shown in Fig.~\ref{FPGA adder}.
%
\begin{eqnarray}\label{Eq:RCA_delay}
	\mu_{RCA}=n\cdot\mu_c
\end{eqnarray}

@@

\subsection{Probabilistic Model of Truncation Error}\label{section_RCA_TruncationError}
For the ease of discussion, we assume that the input to our circuit is a fixed point number scaled to lie in the range $[-1,1)$. In our initial analysis, we assume every bit of each input is uniformly and independently generated. However, this assumption will be relaxed in Section~\ref{Section_Experiments} where the predictions are verified using real image data. The errors at the output are evaluated in terms of the absolute value and the probability of their occurring. These two metrics are combined as the error expectation.

@@

However, since the length of an actual carry chain during execution is dependent upon input patterns, in general, the worst case may occur rarely. To determine when this timing constraint is not met and the size of the error in this case, we expand standard results~\cite{DigitalICDesign} to the following statements, which examine carry generation, propagation and annihilation, as well as the corresponding summation results of a single bit $i$ within an $n$-bit RCA ($0\leq i<n$), according to the relationship between its input patterns $A_i$ and $B_i$:
%
 \begin{itemize}
   \item Carry generation: $A_i=B_i=1$, a new carry chain is generated at bit $i$, and $S_i=C_{i-1}$;
   \item Carry propagation: $A_i\neq B_i$, the carry propagates for this carry chain at bit $i$, and $S_i=0$;
   \item Carry annihilation: $A_i=B_i$ (either $0$ or $1$), the current carry chain annihilates at bit $i$, and $S_i=1$.
 \end{itemize}

@@

Furthermore, it can be seen that the probabilistic models for the magnitude and probability of overclocking errors can be easily generalized for any input distributions, since (\ref{Eq:Output_Error}) and (\ref{Eq:etm}) are not related to the input distribution, while (\ref{Eq:Ptm_RCA}) can be employed as long as the value of $P_{(A_i=B_i=1)}$ and $P_{(A_i\neq B_i)}$ is known.

@@

In our analysis, the stage delay of stage $i$ in the CSA refers to the combination of the $i$-bit carry propagation delay and the delay of multiplexing the carry output. For a CSA with $s$ stages~($s\geqslant 2$), let the stage delay be denoted by $d_{s-1},~\dots,~d_{0}$, where $d_{s-1}$ and $d_{0}$ represent the delay of the most significant and the least significant stages, respectively. We still follow the aforementioned assumption that the critical path delay of the CSA is due to carry propagation and multiplexing the carry output, instead of generating the sum outputs. It should be noted that unlike other stages, the least significant stage of the CSA can be built only by one RCA without multiplexers, since it is directly driven by the carry input. Hence we can obtain the delay of the $i^{th}$ stage as presented in~(\ref{Eq:CSA_SingleStageDelay}), where $\mu_{c}$, $\mu_{mux}$ and $n_i$ denote the delay of $1$-bit carry propagation, the delay of multiplexing and the word-length of the $i^{th}$ stage of the CSA, respectively.
%
\begin{eqnarray}\label{Eq:CSA_SingleStageDelay}
  d_i=\left\{
    \begin{matrix}
      n_i\cdot \mu_{c}+(s-i)\cdot\mu_{mux}, &\textrm{if $i\in\left[1,s-1\right]$}\\
      n_0\cdot \mu_c + (s-1)\cdot\mu_{mux}, & \textrm{if $i=0$}
    \end{matrix}
  \right.
\end{eqnarray}

@@

\begin{figure}[htbp]
        \centering
        \subfigure[\scriptsize The modeled value and the experimental results of the maximum word-length of RCA and CSA.]{
            \includegraphics[width=3.5in]{./Figures/CSA2stage_Model_Timing.eps}
            \label{Fig_CSA2stage_Timing}
        }
        \subfigure[\scriptsize Hardware resource usage for an RCA and a CSA with 2 and 3 stages.]{
            \includegraphics[width=3.5in]{./Figures/CSA3stage_Area.eps}
            \label{Fig_CSA3stage_Area}
        }
    \caption{A comparison between RCA and CSA in terms of the maximum word-length of input signal and the area consumption.}
\end{figure}

Fig.~\ref{Fig_CSA2stage_Timing} also highlights that in comparison to the RCA, the CSA achieves greater word-length when frequency is initially increased. The RCA only outperforms when very high frequency is applied. This is because at low frequencies, although the multiplexer delay limits the word-length of each stage in CSA when compared to RCA, the stage parallelism in CSA enables a greater word-length. However when frequency increases, the multiplexer delay becomes comparable to the delay of the carry chain, and this inhibits the benefits of parallelism.

@@

\subsection{Determination of the Optimum Adder Structure for Given Frequency Requirements}
First, suppose the algorithm designer wished to create a circuit that can run at a given frequency with the minimum achievable output errors and minimum resource usage, i.e. a pair of $\{Area, Frequency\}$ constraint is applied. In this case, the optimum design method is selected according to the following criteria:

\begin{itemize}
  \item Design with the minimum mean value of errors at the output is the optimum design;
  \item If multiple designs achieve the same accuracy, then the design with minimum area is the optimum design;
  \item If both the accuracy and area are identical for multiple designs, they are all treated as the optimum design.
\end{itemize}

For instance, in Fig.~\ref{Fig_CSA_LUT25} we record the mean value of error at outputs with respect to different operating frequencies for both design scenarios when the number of the available LUTs is set to 25. In this graph, we have labeled the optimum adder structure. Note that in this and all the following experiments within this section, in order to apply our models the input data is randomly generated following the uniform distribution.
%
\begin{figure}[t]
  \centering
  \includegraphics[width=3.3in]{./Figures/Error_LUT25.eps}
  \caption{A comparison between two design scenarios when the number of available LUT is 25. The RCA and CSA are investigated in the conventional scenario, while the RCA is explored in the proposed new scenario. The results are obtained from post place-and-route simulations on Xilinx Virtex-6 FPGAs.}
  \vspace{-.5ex}
  \label{Fig_CSA_LUT25}
\end{figure}

We first notice that for all frequency values, the overclocked RCA achieves smaller error expectation than the RCA with truncated operand word-lengths, as predicted by the models for the RCA in Section~\ref{section_RCA_Comparison}. It can also be observed that the CSA cannot be implemented with the original word-length due to the limited area budget and this leads to large truncation errors. Although the CSA with 3 stages outperforms temporarily when frequency is further increased, the overclocked RCA is still the optimum design when high operating frequencies are applied.

>>>>

% Title portion
\title{Imprecise Datapath Design: An Overclocking Approach}
\author{KAN SHI, DAVID BOLAND\\
and GEORGE A. CONSTANTINIDES
\affil{Imperial College London}
}

\begin{abstract}
%As process scaling introduces significant performance variations, much more conservative timing margins will have to be applied to avoid timing violations. Potentially, r
Releasing the tight accuracy requirement would potentially offer greater freedom to create a design with better performance or energy efficiency. In this paper, we compare two different approaches that could trade accuracy for performance. One is the traditional approach where the precision used in the datapath is limited to meet a target latency. The other is a proposed new approach which simply allows the datapath to operate without timing closure. We demonstrate analytically and experimentally that on average our approach obtains either smaller errors for equivalent faster operating frequencies in comparison to the traditional approach. This is because the worst case caused by timing violations only happens rarely, while precision loss results in errors to most data. We also show that for basic arithmetic operations such as addition, applying our approach to the simple building block of ripple carry adders can achieve better accuracy or performance than using faster adder designs to achieve similar latency.

 %operators (i.e. ripple carry adders) can achieve better accuracy or performance than using the traditional method to both original operators and intrinsically faster operators (i.e. carry select adders) which consumes the same silicon area.
\end{abstract}

@@

\section{Introduction}
%With the continuous scaling of CMOS technology to the nanometer regime, circuit performance has increased tremendously over the past decades. However, the drastic variations introduced by higher integration densities is anticipated to be the major obstacle when designing reliable, high performance circuits. Currently, hardware designers tend to employ conservative safety margins to ensure a uniform functionality across a variety of possible working environments. Nevertheless, continuing with this approach could become very costly in the short term future. This is because the highly scaled CMOS devices would inevitably exhibit probabilistic or statistical behavior, meaning that we can no longer expect a uniform circuit performance. Furthermore, covering all possible worst cases would become increasingly difficult, expensive and result in large yield loss because of the variation.

Performance is a critical issue when designing embedded computing systems. Specifically, a large volume of current studies has demonstrated that, significant performance gains can be achieved by FPGA-based accelerators over software designs across a wide range of applications~\cite{UnderwoodFPGAvsCPU,David_MINRES}. However, one of the major factors that limits the performance of these accelerators is that they typically run at much lower clock frequencies than general purpose processors (GPPs) or graphics processing units (GPUs).

In order to boost the operating frequency of a datapath, the standard approaches are either to heavily pipeline the design or to reduce the precision throughout the datapath. For the former method, it should be noted that pipelining will not tend to reduce the latency of the datapath, whilst the latency in terms of clock cycles will actually increase. As a result, this method will not be applicable to many embedded applications, which often have strict latency requirements, or in any datapath containing feedback where C-slow retiming is inappropriate. For the second approach, reducing the datapath precision would reduce the latency at the cost of introducing quantization errors into the design. This has inspired research into exploiting the potential benefits of using the minimum precision necessary to satisfy a certain accuracy or performance specification~\cite{GeorgeDT11}. A brief review of existing approaches in this area will be discussed in Section~\ref{Section_Background_WL}.

Unfortunately, neither of these conventional approaches tends to remove the conservative safety margin, which is introduced to ensure a uniform functionality across a variety of possible working environments. Nevertheless, continuing with employing the guard-bands to cover all worst cases would become increasingly difficult and expensive in the short-term future, because of the continuous scaling of the process technology. In recent years, we have seen a growth of research that explores the potential power or performance benefits that can be obtained when relaxing the guard-bands~\cite{NewBox2004}. The detailed review of this research field will be presented in Section~\ref{Section_Background_Imprecise}.

In this paper, we describe an alternative circuit design methodology when considering trade-offs between accuracy, performance and silicon area. Since it is unlikely to create a completly error-free design as any data representation would introduce quantization errors, we take a radical shift forward by suggesting that for certain applications it is beneficial to move away from the traditional model of creating a conservative design that is guaranteed to avoid timing violations. Instead, it may be preferable to create a design in which timing violations may occur, under the knowledge that they only occur rarely because specific input patterns are required to generate the worst case errors. This paper elaborates on the prior work [xxx], by incorporating silicon area as another evaluation metric. This enables us to find the optimum design choice of the basic arithmetic operators under various design constraints. A summary of the main contributions of this work are as follows:

\begin{itemize}
	\item Detailed modeling methods of probabilistic errors generated in basic arithmetic primitives from two design scenarios to trade accuracy for performance: overclocking the datapath or truncating precisions that used throughout the datpath,
	\item Exploring the accuracy-performance-area trade-offs for key arithmetic operators to illustrate the optimum architecture for a certain arithmetic function under a given requirement of silicon area,
	\item Demonstration of the proposed findings analytically and experimentially from FPGA implementations that allowing rare timing violations to happen leads to a reduction of average errors or an improvement of performance over trancating the datapath precisions to meet timing.
\end{itemize}

@@

\section{Background}\label{Section_Background}
\subsection{Word-length Optimization}\label{Section_Background_WL}
There exists a significant amount of work demonstrating that optimizing the datapath precision would bring a substantial benefits on clock speed, silicon area and power consumption. There are two main research directions in this area. The first discusses methods to analyze the errors caused by the usage of finite precision and compute bounds on the worst-case errors seen at the output of the design. Related techniques include the simulation-based approach~\cite{SungWLop_Simulation} and the analytical-based approach such as interval arithmetic~\cite{moore1966IA}, affine arithmetic~\cite{de2004affine}, Satisfiability-Modulo Theories~\cite{SAT2010} and polynomial representations~\cite{Boland2011TCAD}. A detailed summary of these methods can be found in~\cite{GeorgeDT11}. In the second research field, these tools are utilized to choose the minimum precision within a datapath to satisfy a required accuracy specification~\cite{RoldaoWLop_Simulation}. In this work, two different design scenarios are analyzed. In the conventional scenario, we speculate and reduce the original word-length in the datapath under a given timing constraint while ensuring no timing errors will occur. In the proposed new scenario, we operate the datapath with the original word-length while allowing timing violations sometimes to occur. The truncation errors derived in the former scenario are then compared to the overclocking errors generated in our proposed new scenario.

%we speculate and employ the maximum word-length in the datapath under a given timing constraint while ensuring no timing errors will occur. The truncation errors derived in this process are then compared to the overclocking errors generated in our proposed new scenario.

\subsection{Imprecise Circuit Design Methodology}\label{Section_Background_Imprecise}
It is worth noting that the choice of precision is not the only source of error when designing a datapath. Recently we have seen a growth of parallel streams of research which aim at exploring alternative methods to trade accuracy for design efficiency. This strand of research is motivated by the fact that extra benefits of manufacturing, test, power and timing can be obtained if the tight requirement of absolute correctness is released for devices and interconnect, as pointed out by the International Technology Roadmap for Semiconductors (ITRS) in 2007~\cite{semiconductor2007international}. This topic is expected to be of growing importance in the future, because we will face new design challenges as the technology scales further.

According to the existing literature, one way to operate circuits beyond the deterministic region is to relax the design constraints and the guard-bands that are conventionally used to avoid the worst cases. A series of work known as ``Better-Than-Worst-Case (BTWC) Design'' introduced a universal framework to push the circuit performance to its limits~\cite{BetterThanWS2005}. In general, a BTWC design is composed of cores and checkers. The cores are operated with high performance by eliminating the guard-bands, meanwhile the possible timing errors are diagnosed by the checkers. Optionally the system can be recovered at the observation of errors. As an exemplary design, the Razor project~\cite{Razor2003,Razor2004} was proposed to shave the conservative timing margins by overscaling the supply voltage and clock frequency, while monitoring the output error rates by utilizing a self-checking circuit. This work demonstrated that the benefits brought by removing the safe margin outweigh the cost of monitoring and recovering from errors. For example, $22\%$ or over $30\%$ power consumption can be saved with $\sim0.01\%$ or $\sim1\%$ error rates at the output, respectively. Related work also includes a similar frequency overscaling technique by operating circuits slightly slower than the critical path delay with dedicated checker circuits to ensure that timing errors will not occur~\cite{TEAtime2004}, or developing timing analysis tools that decide the optimum operating frequencies in the non-deterministic region due to process variation~\cite{ProbabilisticSTA}. However, the major problem of the BTWC work is the extra cost introduced by implementing the checker circuits.
%of silicon area due to the implementation of the checker circuits.

Another field of study has shown that errors can be potentially tolerated in various applications, e.g. DSP applications involving human perception. In this case, even greater performance or power benefits can be achieved, as the design overhead of the checkers can be eliminated. Current research in this area focuses on designing probabilistic or imprecise circuits both at a software level and a hardware level. For instance at the programming language level, a tool was proposed~\cite{EnerJ2011Uwash,Truffle2012Uwash} to divide the program into the precise parts and the approximate parts, which are mapped to different hardware with different speed-grades, supply voltages etc., respectively. This technique enables a relatively high service quality, in the meantime energy reduction can be obtained due to approximation. Similar ideas have also been applied on hardware directly. As an example, Palem et~al.~\cite{NonUniformScaling} described a non-uniform voltage scaling technique for the ripple carry adder. In this study, multiple voltage regions are employed for different bits along a carry chain. That is, higher voltage would be applied for computations generating most significant bits, and vice versa. Nevertheless, utilizing several voltage regions within a ripple carry adder will rarely be practical in real situations. Furthermore, the overhead of applying this technique is not addressed.

While the aforementioned literature take advantage of the fact that only specific input patterns could cause timing errors, research on imprecise architectures take this one step further by designing simplified circuits for peformance and/or energy efficiency. For example, Lu et~al. proposed a ``shrinking'' datapath that can be utilized to mimic and speculate the original logic functions~\cite{IntelSpeeding}. Kulkarni et.al. described an underdesigned $2\times2$ multiplier unit, of which the worst case was replaced by a normal case based on the straight-forward Karnaugh-Map analysis~\cite{Undersigned2x2multiplier}. In both cases, reduction of area and power consumption can be achieved with the cost of accuracy. In addition, Gupta et~al. developed approximate adders at the transistor level and compared the energy efficiency of their proposed architectures over truncation of input word-length of conventional structures~\cite{Gupta2013TransCADICS}. However, we should note that using imprecise architectures means correct results would never be obtained for certain input patterns. Another limitation is that the link between the probability of output correctness and energy savings or performance improvements are rarely analyzed in current research. In addition, these techniques cannot be directly applied onto existing hardware platforms such as FPGAs.

In this work, we propose an alternative methodology to remove the limitations of those existing approaches. It is argued in this paper that our method can be easily applied to existing hardware arithmetic operators, and we support this hypothesis analytically by presenting theoretical probabilistic error models, and experimentally by using FPGAs.

\section{Ripple Carry Adder}\label{section_RCA}
\subsection{Adder Structures in FPGAs}
Adders serve as a key building block for arithmetic operations. In general, the ripple carry adder (RCA) is the most straightforward and widely used adder structure. As such, the philosophy of our approach is first exemplified with the analysis of a RCA. We later describe how this methodology can be extended to other arithmetic operators such as CSAs and CCMs, which are commonly used in DSP applications and numerical algorithms, in Section~\ref{Section_CSA} and Section~\ref{section_CCM} respectively.

@@

While the fast carry logic reduces the time of each individual carry-propagation delay, the overall delay of carry-propagation will eventually overwhelm the delay of sum generation of each LUT with increasing operand word-lengths. We perform our initial analysis based on the assumption that the carry propagation delay of each FA is a constant value $\mu_c$, which is a combination of logic delay and routing delay. Hence the critical path delay of the RCA is given by~(\ref{Eq:RCA_delay}), as shown in Fig.~\ref{FPGA adder}.
%
\begin{eqnarray}\label{Eq:RCA_delay}
	\mu_{RCA}=n\cdot\mu_c
\end{eqnarray}

@@

\subsection{Probabilistic Model of Truncation Error}\label{section_RCA_TruncationError}
For ease of discussion, we assume that the input to our circuit is a fixed point number scaled to lie in the range $[-1,1)$. In our initial analysis, we assume every bit of each input is uniformly and independently generated. However, this assumption will be relaxed in Section~\ref{Section_Experiments} where the predictions are verified using real image data. The errors at the output are evaluated in terms of the absolute value and the probability of their occurrence. These two metrics are combined as the error expectation.

@@

However, since the length of an actual carry chain during execution is dependent upon input patterns, in general, the worst case may occur rarely. To determine when this timing constraint is not met and the size of the error in this case, we expand standard results~\cite{DigitalICDesign} to the following statements, which examine carry generation, propagation and annihilation, as well as the corresponding summation results of a single bit $i$ within an $n$-bit RCA ($0\leq i<n$), according to the relationship between its input patterns $A_i$ and $B_i$:
%
 \begin{itemize}
   \item Carry generation: $A_i=B_i=1$, a new carry chain is generated at bit $i$, and $S_i=C_{i-1}$;
   \item Carry propagation: $A_i\neq B_i$, the carry propagates for this carry chain at bit $i$, and $S_i=\bar{C_{i-1}}=0$;
   \item Carry annihilation: $A_i=B_i$ (either $0$ or $1$), the current carry chain annihilates at bit $i$, and $S_i=C_{i-1}=1$.
 \end{itemize}

@@

Furthermore, it can be seen that the probabilistic models for the magnitude and probability of overclocking errors can be easily generalized for any input distributions with mutually independent bits, since (\ref{Eq:Output_Error}) and (\ref{Eq:etm}) are not related to the input distribution, while (\ref{Eq:Ptm_RCA}) can be employed as long as the value of $P_{(A_i=B_i=1)}$ and $P_{(A_i\neq B_i)}$ is known.

@@

In our analysis, the stage delay of the $i^{th}$ stage in the CSA refers to the combination of the $i$-bit carry propagation delay and the delay of multiplexing the carry from the carry input of the $i^{th}$ stage to the carry output of the CSA. For a CSA with $s$ stages~($s\geqslant 2$), let the stage delay be denoted by $d_{s-1},~\dots,~d_{0}$, where $d_{s-1}$ and $d_{0}$ represent the delay of the most significant and the least significant stages, respectively. We still follow the aforementioned assumption that the critical path delay of the CSA is due to carry propagation and multiplexing the carry output, instead of generating the sum outputs. It should be noted that unlike other stages, the least significant stage of the CSA can be built only by one RCA without multiplexers, since it is directly driven by the carry input. Hence we can obtain the delay of the $i^{th}$ stage as presented in~(\ref{Eq:CSA_SingleStageDelay}), where $\mu_{c}$, $\mu_{mux}$ and $n_i$ denote the delay of $1$-bit carry propagation, the delay of multiplexing and the word-length of the $i^{th}$ stage of the CSA, respectively.
%
\begin{eqnarray}\label{Eq:CSA_SingleStageDelay}
  d_i=\left\{
    \begin{matrix}
      n_i\cdot \mu_{c}+(s-i)\cdot\mu_{mux}, &\textrm{if $i\in\left[1,s-1\right]$}\\
      n_0\cdot \mu_c + (s-1)\cdot\mu_{mux}, & \textrm{if $i=0$}
    \end{matrix}
  \right.
\end{eqnarray}

@@

\begin{figure}[htbp]
        \centering
        \subfigure[\scriptsize The modeled value and the experimental results of the maximum word-length of RCA and CSA, with respect to various frequency requirements.]{
            \includegraphics[width=3.5in]{./Figures/CSA2stage_Model_Timing.eps}
            \label{Fig_CSA2stage_Timing}
        }
        \subfigure[\scriptsize Hardware resource usage for an RCA and a CSA with 2 and 3 stages.]{
            \includegraphics[width=3.5in]{./Figures/CSA3stage_Area.eps}
            \label{Fig_CSA3stage_Area}
        }
    \caption{A comparison between RCA and CSA in terms of the maximum word-length of input signal and the area consumption.}
\end{figure}

Fig.~\ref{Fig_CSA2stage_Timing} also highlights that for a relatively relexed frequency cosntraint, the CSA achieves greater word-length than the RCA. This is because the stage parallelism in the CSA enables a greater overall word-length, even though the multiplexer delay limits the word-length of each stage in the CSA when compared to the RCA. When more stringent latency requirement is applied, however, the word-length of each stage is largely limited such that the multiplexer delay becomes comparable to the delay of the carry chain, and this inhibits the benefits of parallelism.

%in comparison to the RCA, the CSA achieves greater word-length for relatively relexed frequency constraints. The RCA only outperforms when very high frequency requirement is applied. This is because at low frequencies, although the multiplexer delay limits the word-length of each stage in CSA when compared to RCA, the stage parallelism in CSA enables a greater word-length. However when more stringent latency requirement is applied, the word-length of each stage in CSA is limited. Hence the multiplexer delay becomes comparable to the delay of the carry chain, and this inhibits the benefits of parallelism.

@@

\subsection{Determination of the Optimum Adder Structure for Given Frequency Requirements}
First, suppose the algorithm designer wished to create a circuit that can run at a given frequency in a given resource constraint while achieving the minimum possible output error. A decision must then be taken over which adder structure achieves this minimum, and whether the structure should be overclocked

% with the minimum achievable output errors and minimum resource usage, i.e. a pair of $\{Area, Frequency\}$ constraint is applied. In this case, the optimum design method is selected according to the following criteria:

%\begin{itemize}
%  \item Design with the minimum mean value of errors at the output is the optimum design;
%  \item If multiple designs achieve the same accuracy, then the design with minimum area is the optimum design;
%  \item If both the accuracy and area are identical for multiple designs, they are all treated as the optimum design.
%\end{itemize}

As an example slice through the design space, in Fig.~\ref{Fig_CSA_LUT25} we record the mean value of error at outputs with respect to different operating frequencies for both design scenarios when the number of the available LUTs is set to 25. In this graph, we have labeled the optimum adder structure. Note that in this and all the following experiments within this section, in order to apply our models the input data is randomly generated following the uniform distribution.
%
\begin{figure}[t]
  \centering
  \includegraphics[width=3.3in]{./Figures/Error_LUT25.eps}
  \caption{A comparison between two design scenarios when the number of available LUT is 25. The RCA and CSA are investigated in the conventional scenario, while the RCA is explored in the proposed new scenario. The results are obtained from post place-and-route simulations on Xilinx Virtex-6 FPGAs.}
  \vspace{-.5ex}
  \label{Fig_CSA_LUT25}
\end{figure}

We first notice that for all frequency values, the overclocked RCA achieves no larger error expectation than the RCA with truncated operand word-lengths, as predicted by the models for the RCA in Section~\ref{section_RCA_Comparison}. It can also be observed that the CSA cannot be implemented with the original word-length due to the limited area budget and this leads to large truncation errors. Although the CSA with 3 stages is best for some frequencies, the overclocked RCA is still the optimum design when high operating frequencies are applied.
